\documentclass[11pt]{article}
\usepackage[paperwidth=8.5in,paperheight=11in,margin=1in]{geometry}
\usepackage{xr}
\usepackage{amsmath}
\externaldocument{paper}

% For section headers starting with S
\renewcommand{\thesection}{S.\arabic{section}}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}

% Equation numbering in supplementary style
\makeatletter
\def\tagform@#1{\maketag@@@{(S\ignorespaces#1\unskip\@@italiccorr)}}
\makeatother

\newcommand{\myfigurename}{Figure}
\newcommand{\myref}[1]{\ref{#1}}
\newcommand{\suppref}[1]{S\ref{#1}}
\usepackage{amsmath,amssymb}
\DeclareMathOperator{\EX}{\mathrm{E}}
\DeclareMathOperator{\Var}{\mathrm{Var}}
\DeclareMathOperator{\Cov}{\mathrm{Cov}}
\usepackage{graphicx}

\title{Supporting Information}
\author{}

\begin{document}
\maketitle

\section{Methods}

\subsection{The EAGGL model}

EAGGL (Enrichment Analysis Gene Grouping and LLMs) is the latent-structure stage that follows PIGEAN. Whereas PIGEAN estimates the probability that each gene is relevant to a trait and the degree to which each gene annotation predicts trait-relevance, the goal of EAGGL is to decompose those trait-associated gene and annotation patterns into a small number of latent mechanisms.

Let genes be indexed by $i = 1 \ldots N$, gene annotations or gene sets by $j = 1 \ldots M$, phenotypes by $t = 1 \ldots T$, and latent mechanisms by $k = 1 \ldots K$. Let $X_{ij} \ge 0$ denote the value of annotation $j$ for gene $i$. In the simplest setting, $X_{ij}$ is binary gene-set membership, although weighted annotations are also permitted. Let $p_i^{(u)} \in [0,1]$ denote the gene-level support for gene $i$ for anchor user $u$, and let $q_j^{(u)} \in [0,1]$ denote the corresponding annotation-level or gene-set-level support for annotation $j$ for anchor user $u$. The index $u = 1 \ldots U$ is a generic anchor index; depending on the workflow, it may refer to the default phenotype, a supplied positive-control pseudo-phenotype, one or more anchor phenotypes, one or more anchor genes, or an anchor gene set.

The basic biological assumption is that trait-relevant genes and trait-relevant annotations co-occur because they load on a small number of shared mechanisms. In the gene-by-annotation workflow family, we therefore posit nonnegative gene weights $W_{ik}$ and nonnegative annotation weights $H_{jk}$ such that
\begin{equation}
X_{ij} \approx \sum_{k=1}^K W_{ik} H_{jk} + \epsilon_{ij},
\label{eq:eaggl_basic_factor}
\end{equation}
where $\epsilon_{ij}$ is an error term. A factor $k$ is interpreted as a latent mechanism when a coherent set of genes has large $W_{ik}$ values and a coherent set of annotations has large $H_{jk}$ values.

The entries of $X$ are not equally informative for a given trait. For example, an annotation that is itself strongly implicated by PIGEAN should contribute more strongly to inference than an annotation with no trait support; likewise, genes with high trait support should contribute more than genes with no support. EAGGL therefore fits the factorization under a weighted Gaussian likelihood:
\begin{equation}
\Pr(X \mid W, H, \Omega)
\propto
\exp\left(
-\frac{1}{2}
\sum_{i=1}^N \sum_{j=1}^M
\Omega_{ij}
\left(
X_{ij} - \sum_{k=1}^K W_{ik} H_{jk}
\right)^2
\right),
\label{eq:eaggl_weighted_likelihood}
\end{equation}
where $\Omega_{ij} \ge 0$ is a cell-specific weight. In the simplest single-trait setting, one can think of $\Omega_{ij}$ as proportional to $p_i q_j$, where $p_i$ is gene support and $q_j$ is annotation support. More generally, when several anchor users are present, EAGGL uses the additive form
\begin{equation}
\Omega_{ij}
=
\sum_{u=1}^U p_i^{(u)} q_j^{(u)}.
\label{eq:eaggl_weight_matrix}
\end{equation}
This gives higher weight to entries supported simultaneously by the relevant genes and the relevant annotations for the chosen anchor set.

Although EAGGL is often motivated heuristically by factorizing a matrix of the form $X_{ij} p_i q_j$, the weighted likelihood in Eq.~\eqref{eq:eaggl_weighted_likelihood} is the more useful mathematical form. It makes explicit that EAGGL seeks a factorization of the underlying annotation matrix while emphasizing entries most relevant to the trait or anchor set. This also avoids repeatedly materializing a different dense weighted matrix for every anchor user.

\subsection{Fitting the model}

Because the factor matrices are latent, we infer them from the observed nonnegative matrix and the support-derived weights. The factorization is regularized by automatic relevance determination (ARD), which allows the effective number of factors to be learned from the data rather than fixed in advance.

Specifically, for each factor $k$ we introduce a nonnegative relevance scale $\lambda_k$ and use the prior
\begin{equation}
W_{ik} \mid \lambda_k \sim \text{HalfNormal}\left(0, \lambda_k / \phi \right),
\qquad
H_{jk} \mid \lambda_k \sim \text{HalfNormal}\left(0, \lambda_k / \phi \right),
\label{eq:eaggl_halfnormal_prior}
\end{equation}
together with a factor-specific ARD prior
\begin{equation}
\lambda_k \sim \text{InvGamma}(a_0, b_0).
\label{eq:eaggl_ard_prior}
\end{equation}
Here $\phi$ controls the baseline sparsity of the factor loadings, while $a_0$ and $b_0$ control the tendency of entire factors to be shrunk away. Small values of $\lambda_k$ correspond to factors that explain little of the observed matrix after accounting for the regularization, and are therefore removed.

Up to additive constants, maximizing the posterior is equivalent to minimizing
\begin{equation}
\mathcal{L}(W,H,\lambda)
=
\frac{1}{2}
\sum_{i=1}^N \sum_{j=1}^M
\Omega_{ij}
\left(
X_{ij} - \sum_{k=1}^K W_{ik} H_{jk}
\right)^2
+
\frac{\phi}{2}
\sum_{k=1}^K
\frac{
\|W_{\cdot k}\|_2^2 + \|H_{\cdot k}\|_2^2 + 2b_0
}{
\lambda_k
}
+
C
\sum_{k=1}^K
\log \lambda_k,
\label{eq:eaggl_objective}
\end{equation}
where
\begin{equation}
C = \frac{N + M}{2} + a_0 + 1.
\label{eq:eaggl_C}
\end{equation}
Thus, EAGGL uses the same basic intuition as other Bayesian nonnegative matrix factorization procedures with ARD: the reconstruction term encourages faithful explanation of the observed matrix, while the ARD penalty favors retaining only factors whose contribution is large enough to overcome whole-factor shrinkage.

The number of retained mechanisms is therefore not simply the user-specified maximum candidate rank. Rather, EAGGL begins from a candidate rank $K$ and then discards factors with very small posterior relevance scales or nearly zero total loading. In this way, the effective number of mechanisms is learned adaptively.

\subsection{Single-trait gene-by-annotation factorization}

The default EAGGL workflow operates on the gene-by-annotation matrix $X$ together with a single trait-specific gene support vector $p_i$ and annotation support vector $q_j$. In this case $U=1$ and the weight matrix simplifies to
\begin{equation}
\Omega_{ij} = p_i q_j.
\label{eq:eaggl_single_trait_weights}
\end{equation}
The fitted factor matrices $W$ and $H$ may be interpreted as:
\begin{itemize}
\item $W_{ik}$: the degree to which gene $i$ participates in mechanism $k$;
\item $H_{jk}$: the degree to which annotation $j$ captures mechanism $k$.
\end{itemize}
This is the primary EAGGL model and corresponds to the scientific setting in which one seeks to decompose the trait-associated gene-by-annotation landscape into a small number of mechanisms.

A useful special case is when the input support is not obtained from a standard PIGEAN trait run, but instead from a manually supplied gene list. In that standalone workflow, EAGGL first uses the loaded gene universe from $X$ as the enrichment background and computes a hypergeometric enrichment $p$-value for every annotation or gene set. If $M$ is the number of genes in the loaded universe, $n$ is the number of supplied genes present in that universe, $N_j$ is the size of annotation $j$, and $k_j$ is the overlap between annotation $j$ and the supplied gene list, then
\begin{equation}
p_j^{(\mathrm{hyper})}
=
\Pr\!\left(
\mathrm{Hypergeom}(M, n, N_j) \ge k_j
\right).
\label{eq:eaggl_gene_list_hypergeom}
\end{equation}
After Benjamini--Hochberg correction, annotations with $q_j \le q_{\max}$ are retained. Their annotation-level support is then defined by
\begin{equation}
q_j^{(\mathrm{standalone})}
=
\frac{-\log p_j^{(\mathrm{hyper})}}{\sqrt{N_j}},
\label{eq:eaggl_gene_list_weights}
\end{equation}
while the gene-level support vector is binary:
\[
p_i^{(\mathrm{standalone})}
=
\begin{cases}
1 & \text{if gene } i \text{ is in the supplied list,} \\
0 & \text{otherwise.}
\end{cases}
\]
All genes appearing in any retained annotation are then kept in the final factoring matrix. The factorization itself is unchanged; only the way the gene-level and annotation-level support vectors are constructed differs from the standard PIGEAN-driven path.

\subsection{Phenotype-anchored extensions of the gene-by-annotation model}

In some applications, one wishes to condition factorization on several anchor phenotypes rather than on a single default trait. Let $P_{it}$ denote a nonnegative gene-by-phenotype support matrix and $Q_{jt}$ denote a nonnegative annotation-by-phenotype support matrix derived from phenotype-resolved statistics. If the anchor phenotype set is $A \subseteq \{1,\ldots,T\}$, then each $t \in A$ defines an anchor user, with
\[
p_i^{(u)} = P_{it},
\qquad
q_j^{(u)} = Q_{jt}.
\]
The weight matrix again takes the additive form of Eq.~\eqref{eq:eaggl_weight_matrix}:
\begin{equation}
\Omega_{ij}^{(A)}
=
\sum_{t \in A}
P_{it} Q_{jt}.
\label{eq:eaggl_pheno_anchor_weights}
\end{equation}
The factorization itself remains the same as Eq.~\eqref{eq:eaggl_basic_factor}. The difference is that the model is now conditioned on a selected phenotype set rather than on a single default trait.

When the goal is to summarize relevance to any phenotype in a large panel rather than to a manually specified subset, the anchor set is first collapsed to a single aggregate user and factorization proceeds under the same weighted model. Thus, the difference between multi-phenotype anchoring and any-phenotype anchoring lies in whether the phenotype-resolved support vectors are retained separately or combined before fitting; the underlying factor model is unchanged.

\subsection{Gene-anchored and gene-set-anchored gene-set by phenotype factorization}

The second major EAGGL workflow family factors a nonnegative gene-set-by-phenotype matrix rather than the original gene-by-annotation matrix. Let $Y_{jt} \ge 0$ denote a matrix whose rows are gene sets and whose columns are phenotypes, typically derived from phenotype-resolved PIGEAN statistics. EAGGL then posits
\begin{equation}
Y_{jt}
\approx
\sum_{k=1}^K H_{jk} Z_{tk} + \epsilon_{jt},
\label{eq:eaggl_gene_set_pheno_factor}
\end{equation}
where $H_{jk}$ is again the loading of annotation $j$ on factor $k$, while $Z_{tk}$ is now the phenotype loading of factor $k$.

As above, the factorization is fitted under a weighted Gaussian likelihood. Let $q_j^{(u)}$ denote gene-set support and let $r_t^{(u)}$ denote phenotype support induced by anchor user $u$. Then
\begin{equation}
\Omega_{jt}
=
\sum_{u=1}^U q_j^{(u)} r_t^{(u)}.
\label{eq:eaggl_gene_anchor_weights}
\end{equation}
The anchor users are defined differently for the different workflows in this family:
\begin{itemize}
\item In single-gene anchoring and multi-gene anchoring, $u$ indexes one or more selected anchor genes.
\item In any-gene anchoring, the selected anchor genes are first collapsed into a single aggregate user.
\item In gene-set anchoring, the user is a supplied anchor gene set that induces phenotype support through the gene-PheWAS statistics.
\end{itemize}
Thus, the gene-anchored workflows and the phenotype-anchored workflows are not different factorization algorithms. They are different constructions of the matrix and weights supplied to the same ARD-regularized nonnegative factor model.

\subsection{Factor relevance and any-anchor relevance}

After fitting, EAGGL reports factor-level relevance scores that summarize how strongly each factor is supported by the selected trait or anchor set. These are downstream summaries derived from the fitted factor basis and are not themselves part of the factorization objective.

For the gene-by-annotation workflow family, let $W \in \mathbb{R}_+^{N \times K}$ denote the fitted gene-by-factor matrix, and let $p^{(u)} \in [0,1]^N$ denote the gene-level support vector for anchor user $u$. The anchor-specific relevance of the factors to user $u$ is defined by bounded nonnegative regression:
\begin{equation}
r^{(u)}
=
\arg\min_{0 \le r \le 1}
\left\|
p^{(u)} - W r
\right\|_2^2.
\label{eq:eaggl_factor_relevance_gene}
\end{equation}

For the gene-set-by-phenotype workflow family, let $Z \in \mathbb{R}_+^{T \times K}$ denote the fitted phenotype-by-factor matrix, and let $p^{(u)} \in [0,1]^T$ denote the phenotype support vector for anchor user $u$. The factor relevance is then defined analogously:
\begin{equation}
r^{(u)}
=
\arg\min_{0 \le r \le 1}
\left\|
p^{(u)} - Z r
\right\|_2^2.
\label{eq:eaggl_factor_relevance_pheno}
\end{equation}

When more than one anchor user is present, EAGGL summarizes the probability-like relevance of factor $k$ to any selected anchor by a noisy-OR aggregation:
\begin{equation}
r_k^{\mathrm{any}}
=
1 -
\prod_{u=1}^U
\left(
1 - r_k^{(u)}
\right).
\label{eq:eaggl_factor_any_relevance}
\end{equation}
This has the intended semantics that $r_k^{\mathrm{any}}$ is the probability-like estimate that factor $k$ is relevant to at least one selected anchor. In the single-user case, Eq.~\eqref{eq:eaggl_factor_any_relevance} reduces to the anchor-specific relevance itself.

\section{Phenotype capture and factor-specific phenotype enrichment}
\label{sec:postfactor_revised}

After fitting the main factorization, EAGGL can relate latent factors to additional human phenotypes in two conceptually distinct ways. The first is a descriptive \emph{capture} step, which estimates how strongly each phenotype's retained high-confidence support profile is represented by each factor. The second is an inferential \emph{enrichment} step, which asks whether a phenotype is more concentrated in one factor than expected after accounting for the anchor trait and optional covariates. These two quantities answer different scientific questions and are therefore not expected to coincide exactly.

\subsection{Phenotype capture by compositional projection of thresholded profiles}

The external gene-by-phenotype support table can be very large. In practice, EAGGL therefore consumes a sparse thresholded representation that retains only gene-phenotype entries whose combined support exceeds a cutoff $c_0$. The phenotype profile used for post-factor annotation is therefore not the full unthresholded phenotype surface. It is the retained high-confidence thresholded profile.

Let $b$ denote a feature basis shared by the factors and a phenotype-specific support vector. Depending on the workflow, this basis may be genes, gene sets, or phenotypes already present in the factored matrix. Let
\[
B^{(b)} \in \mathbb{R}_+^{F_b \times K}
\]
denote the factor basis on that feature space, where $F_b$ is the number of features in basis $b$, and let
\[
s_t^{(b)} \in \mathbb{R}_+^{F_b}
\]
denote the thresholded phenotype profile for phenotype $t$ on that same basis.

By default, EAGGL uses a \emph{weighted thresholded} profile: retained combined-support values are used for entries above the cutoff and zero otherwise. As an expert sensitivity analysis, one may instead use a \emph{binary thresholded} profile, in which retained entries are coded as $1$ and absent entries as $0$ before normalization.

We decompose each thresholded phenotype profile into an overall retained strength and a compositional shape. Define the retained phenotype strength
\begin{equation}
a_t^{(b)} = \sum_{f=1}^{F_b} s_{tf}^{(b)},
\label{eq:eaggl_pheno_strength}
\end{equation}
and the corresponding compositional phenotype profile
\begin{equation}
\tilde{s}_t^{(b)} = \frac{s_t^{(b)}}{a_t^{(b)} + \epsilon},
\label{eq:eaggl_pheno_profile_normalized}
\end{equation}
where $\epsilon>0$ is a small constant to avoid division by zero. Likewise, each factor basis vector is normalized to unit $L_1$ mass,
\begin{equation}
\tilde{B}^{(b)}_{\bullet k} = \frac{B^{(b)}_{\bullet k}}{\sum_{f=1}^{F_b} B^{(b)}_{fk} + \epsilon}.
\label{eq:eaggl_factor_basis_normalized}
\end{equation}

The phenotype capture weights are then defined by constrained nonnegative projection:
\begin{equation}
c_t^{(b)}
=
\arg\min_{c \ge 0,\ \mathbf{1}^\top c \le 1}
\left\|
\tilde{s}_t^{(b)} - \tilde{B}^{(b)} c
\right\|_2^2.
\label{eq:eaggl_phenotype_capture}
\end{equation}
The coefficient $c_{tk}^{(b)}$ is interpreted as the degree to which phenotype $t$ is captured by factor $k$ on basis $b$. In contrast to a projection on raw support mass, Eq.~\eqref{eq:eaggl_phenotype_capture} separates retained phenotype breadth from phenotype-specific structure. Broad phenotypes can therefore have large retained strength $a_t^{(b)}$ without automatically receiving high capture weights for every factor.

\paragraph{Interpretation.}
Phenotype capture is a descriptive quantity. It answers the question:
\begin{quote}
``How is the retained high-confidence support profile of phenotype $t$ represented as a mixture of latent mechanisms?''
\end{quote}
It is not a p-value, and it should not be interpreted as formal evidence that phenotype $t$ is uniquely associated with factor $k$.

\subsection{Choice of projection basis across workflow families}

For the default gene $\times$ gene-set workflow family (F1--F5), the preferred basis for phenotype capture is the gene basis. In that case, $B^{(b)}$ is the gene-by-factor loading matrix and $s_t^{(b)}$ is the thresholded phenotype-specific gene-support vector. If gene-level support is unavailable or gene-level factor loadings are absent, EAGGL falls back to the gene-set basis, in which $B^{(b)}$ is the gene-set-by-factor loading matrix and $s_t^{(b)}$ is the thresholded phenotype-specific gene-set support vector.

For the gene-set $\times$ phenotype workflow family (F6--F9), phenotypes that are already present in the factored matrix have native phenotype loadings. These native loadings are interpreted as phenotype capture weights after appropriate normalization. For external phenotypes that were not part of the factored matrix, EAGGL projects them on the gene-set basis using Eq.~\eqref{eq:eaggl_phenotype_capture}.

Thus, the phenotype-capture definition is common across workflows: EAGGL always projects a thresholded phenotype support profile onto the factor basis defined on the most natural shared feature space.

\subsection{Factor-specific phenotype enrichment with thresholded binary outcomes}

A second, distinct summary asks whether phenotype $t$ is selectively enriched in genes belonging to factor $k$, after accounting for the main anchor trait and optional covariates. Because the standard phenotype input is thresholded, the default factor-specific enrichment analysis is based on a binary thresholded outcome rather than on truncated continuous support values. Let
\[
h_{it} = \mathbb{I}\{ \text{combined phenotype support for gene } i \text{ and phenotype } t \text{ exceeds } c_0 \}
\]
denote the resulting thresholded phenotype-hit indicator. Let $W_{ik}$ denote the gene-membership weight of factor $k$ for gene $i$, let $a_i^{\mathrm{direct}}$ denote the direct support of the main anchor trait for gene $i$, and let $C_i$ denote optional gene-level covariates. The default model fits, for each phenotype $t$ and factor $k$,
\begin{equation}
h_{it}
=
\alpha_{tk}
+
\beta_{tk} W_{ik}
+
\gamma_t a_i^{\mathrm{direct}}
+
\eta_t^\top C_i
+
\varepsilon_{it}.
\label{eq:eaggl_factor_pheno_regression}
\end{equation}
This regression is fit one factor at a time, i.e. as a marginal anchor-adjusted binary model. The coefficient $\beta_{tk}$ is interpreted as the factor-specific enrichment of phenotype $t$ beyond what is already captured by the anchor trait's direct support.

\paragraph{Interpretation.}
Unlike the capture weights in Eq.~\eqref{eq:eaggl_phenotype_capture}, the coefficients in Eq.~\eqref{eq:eaggl_factor_pheno_regression} are incremental effects. They answer the question:
\begin{quote}
``Is phenotype $t$ enriched in factor $k$ beyond the anchor trait's direct gene support?''
\end{quote}
This is an inferential or hypothesis-testing quantity rather than a descriptive loading.

\paragraph{Recommended default and scope.}
The binary thresholded model is the default because it is defensible under sparse thresholded phenotype input and easier to interpret than a joint all-factor fit. EAGGL reports robust uncertainty by default, typically using HC3-type standard errors or chromosome-clustered uncertainty when location information is available. The resulting p-values are still approximate, because the thresholded phenotype-hit process is derived from a combined-support quantity that already reflects gene-set-mediated indirect support.

An expert may instead fit: (i) a marginal binary model without an anchor covariate; (ii) a joint all-factor binary model with direct anchor adjustment; or (iii) legacy continuous direct/combined regressions. Using combined anchor support as the covariate is also an expert-only approximation, because then both the outcome and the covariate inherit related gene-set-mediated structure.

When multiple factor-specific enrichment models are requested in one run, EAGGL appends them into a single output table and tags each row with the model name, factor-model scope, outcome surface, and anchor-adjustment setting so that marginal, joint, and legacy analyses are not conflated downstream.

\subsection{Relationship between phenotype capture and factor-specific enrichment}

The two phenotype summaries serve complementary roles:
\begin{itemize}
\item \textbf{Phenotype capture} gives a bounded, compositional summary of how well the retained high-confidence phenotype profile is represented by each factor. This is the primary annotation layer for interpreting factors with external phenotypes.
\item \textbf{Factor-specific enrichment regression} tests whether that phenotype profile is unusually concentrated in one factor beyond what is already captured by the anchor trait and optional covariates. This is a secondary, expert-oriented summary.
\end{itemize}

Consequently, the main phenotype labels and top-phenotype annotations for a factor should be based on phenotype capture weights, whereas regression coefficients and associated significance measures should be interpreted as downstream enrichment diagnostics rather than as the primary phenotype annotation of the factor.

\subsection{Repeated restarts, consensus factorization, and structural selection of $\phi$}

The EAGGL objective is non-convex, and therefore different random initializations can converge to different local optima. For this reason, EAGGL is typically fit under repeated random restarts. The simplest use of repeated restarts is to retain the fitted solution with the best objective value. A more stringent alternative is consensus factorization, in which factors are first matched across runs and then aggregated only if they are reproduced consistently across restarts.

Let $H^{(r)}$ denote the annotation loading matrix from restart $r$ after ARD thresholding. In consensus mode, EAGGL normalizes the factor columns, aligns factors across restarts by maximum total similarity, and aggregates matched factors only when they appear in a sufficient fraction of runs. Consensus is therefore defined at the component level rather than by averaging arbitrary column order.

The hyperparameter $\phi$ also affects the effective number and sparsity of retained factors, even though ARD is responsible for pruning irrelevant components. EAGGL can therefore treat $\phi$ as a structural model-selection parameter rather than as a fixed constant. The guiding principle is to select the smallest value of $\phi$ that yields a factorization with many retained mechanisms but without unstable or near-duplicate factors.

Operationally, each candidate value of $\phi$ is evaluated across one or more random restarts, and the candidate is summarized by: (a) the modal retained factor count; (b) reconstruction error; (c) agreement of matched factors across runs when multiple restarts are requested; and (d) within-run redundancy of the retained factors. In the current implementation, redundancy is measured primarily on normalized gene loading vectors, with fallback to gene-set or phenotype loading vectors only when gene loadings are unavailable. The overlap score is weighted Jaccard:
\begin{equation}
J_w(u,v)
=
\frac{\sum_{\ell} \min(u_{\ell}, v_{\ell})}{\sum_{\ell} \max(u_{\ell}, v_{\ell})}.
\label{eq:eaggl_weighted_jaccard}
\end{equation}
The selected value of $\phi$ is the smallest value that preserves an acceptably non-redundant and restart-stable solution without materially degrading reconstruction fit. In the current defaults, the redundancy gate is set to weighted Jaccard $0.5$ on the gene-loading basis, corresponding roughly to a requirement that no retained factor share much more than half of its gene-loading mass with another retained factor. This procedure is best understood as structural model selection rather than as held-out cross-validation.
In the implementation, the selected $\phi$, the candidate-level search diagnostics, and the thresholds used for the search are written to the run log and parameter output so that the final reported factorization is auditable after the fact. For very large annotation collections, the implementation evaluates candidate $\phi$ values on a sentinel subset of relatively uncorrelated gene sets and may also use a smaller candidate iteration budget, then reruns the final reported factorization on the full retained matrix with the selected $\phi$. The current defaults use a sentinel size of 1000 representative gene sets and one restart per candidate, with larger restart counts available as an expert stability check. This shortcut changes only the cost of the search procedure; it does not change the final fitting target once $\phi$ has been chosen.

\subsection{Optimizations for scaling EAGGL to large phenotype panels}

Although the model is defined in terms of weighted nonnegative factorization, a direct implementation that explicitly constructs separate dense matrices for every anchor user would be inefficient. EAGGL therefore uses the weighted objective of Eq.~\eqref{eq:eaggl_weighted_likelihood} directly, rather than materializing large user-specific tensors. This preserves the model while substantially reducing memory use.

Similarly, phenotype panels and annotation collections can be large enough that a naive factorization would be dominated by nearly null rows or columns. EAGGL therefore permits filtering and pruning of genes, annotations, or phenotypes before fitting. These operations do not change the form of the statistical model; they change only the subset of the matrix to which the model is applied.

Finally, after fitting, very small factor loadings may be thresholded to zero. This encourages a sparse and more interpretable representation, stabilizes downstream summaries such as top genes and top annotations, and improves comparability across repeated restarts.

\subsection{Interpretation and labeling of latent mechanisms}

The outputs of EAGGL are the fitted factor loadings over genes, annotations, and, when applicable, phenotypes. These loadings constitute the statistical model output. Human-readable labels for factors are downstream summaries rather than part of the factorization model itself.

If $\mathcal{S}_k$, $\mathcal{G}_k$, and $\mathcal{P}_k$ denote the highest-weight annotations, genes, and phenotypes for factor $k$, then a label may be viewed abstractly as a post hoc summary function
\[
L_k = \mathcal{L}(\mathcal{S}_k, \mathcal{G}_k, \mathcal{P}_k).
\]
Such labels may be generated deterministically from the top terms or by a separate language-model-based summarization step. In either case, the labeling procedure is downstream of the core statistical model and should be interpreted as an aid to biological interpretation rather than as part of the inference procedure itself.

\end{document}
